---
title: "Simulate binary logistic data for prob responses"
author: "Andrew Olson"
date: "2025-01-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

For an analysis of serial position information using data that
lists the segments from targets that are preserved in responses.
These data can be inherently ambiguous:e.g. plibebas > plibas
Is the position for the "b" in the response preserved from position
4 or position 7 from the target?  Our approach has been to assign 0.5 to
each ambiguous position in this case.  Based on how binary logistic regression works, this should be the same as simulating a number of responses with the preserved value 1 in position 4 half of the time (position 7 = 0) and the preserved value 1 in position 7 half of the time (position 4 = 0).  A simulation should be able to show that this is correct and the fit is the same.

In sum: The question is whether assigning a value
of 0.5 (not 0/1) to two positions when there is ambiguity is
the same as assigning 1 to one position half of the time and 1
to the other position half of the time.

```{r}
# first create a logistic curve to verify our method for that
x_values <- seq(0,8,length.out=500)
```

```{r}
mean_x <- mean(x_values)
# value of the probability function based on x
prob <- 1/(1+ exp(-(x_values-mean_x)))
```

```{r}
# plot probability as a function of x
p_func_df <- data.frame(x=x_values, prob=prob)
pplt <- ggplot(p_func_df,aes(x=x,y=prob)) + geom_line()
pplt
```
```{r}
# simulate data -- make most of the values 0/1 
# and a smaller number fractional
num_zero_one <- 500
num_fractional_values <- 20 # we will set success=0.5 for these below
# our simulated fractional values
# must be divisible by four because we 
# divide between position 4 and 6 below
# and each position must have 1/2 successes in our binary data below
# and the binary data will have 1/2 the position 4/6 values
num_values <- num_zero_one + num_fractional_values
x_values <- seq(0,10,length.out=num_zero_one)
mean_x <- mean(x_values)
frac_x1 <- rep(4,num_fractional_values/2) # position 4
frac_x2 <- rep(6,num_fractional_values/2) # position 6
prob_first_x <- rep(0.5,num_fractional_values)
frac_data <- data.frame(x=c(frac_x1,frac_x2,x_values))
initial_probs <- rep(0.5,num_fractional_values)
probs_from_logistic <- 1/(1 + exp(-(frac_data$x[num_fractional_values+1:num_zero_one]-mean_x)))
frac_data$prob <- c(initial_probs,probs_from_logistic)
frac_data$random <- runif(num_values)
frac_data$success <- ifelse(frac_data$prob > frac_data$random,1,0)
# replace 0/1 for first values with prob
frac_data$success[1:num_fractional_values] <-
  frac_data$prob[1:num_fractional_values]
frac_data$weights <- c(rep(1,num_fractional_values),rep(1,num_zero_one))
frac_data
```

```{r}
prob_function_data <- frac_data[num_fractional_values+1:num_zero_one,]
dplt <- ggplot(frac_data,aes(x=x,y=success)) + geom_point() +
  geom_line(data=prob_function_data, aes(x=x,y=prob),color="blue")
dplt
```

```{r}
Mfrac <- glm(success ~ x,data=frac_data,family="binomial",weights=weights
             )
Mfrac
```

```{r}
# Now we model the 20 position 4 and 6 values using just 0/1 and no fractional values
# As above, 10 values are in position 4 and 10 in position 6
# half of the position 4 values are 1 and half are 0
# half of the position 6 values are 1 and half are 0
# This simulates an ambiguity where we assign 1 to position 4 and 0 to position 6
# (so putting the preserved item in position 4) half of the time
# and do the opposite, to put the preserved item in position 6, the other half of the time
num_fractional_zero_one <- num_fractional_values
num_values <- num_zero_one + num_fractional_zero_one
frac_x <- c(rep(4,num_fractional_zero_one/2),
            rep(6,num_fractional_zero_one/2))
binary_data <- data.frame(x=c(frac_x,
     frac_data$x[num_fractional_values+1:num_zero_one]))
binary_data$prob <- c(rep(0.5,num_fractional_zero_one),
     frac_data$prob[num_fractional_values+1:num_zero_one])
binary_data$success <- c(rep(0,num_fractional_zero_one/4), # each pos should have 1/2 success 1/2 fail
                         rep(1,num_fractional_zero_one/4),
                         rep(0,num_fractional_zero_one/4),
                         rep(1,num_fractional_zero_one/4),
     frac_data$success[num_fractional_values+1:num_zero_one])
binary_data$random <- c(rep(-1,num_fractional_zero_one),
     frac_data$random[num_fractional_values+1:num_zero_one])
binary_data
```

```{r}
Mbinary <- glm(success ~ x,data=binary_data,family="binomial")
Mbinary
```
